---
title: "wine_report"
output: html_document
date: '2022-06-16'
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../"
    )
  })
---

```{r  results = FALSE, message=FALSE, warning=FALSE}
#loading required packages
library(tidyverse)
library(dplyr)
library(janitor)
library(skimr)
library(GGally)
library(gridExtra)
library(here)
library(DT)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#reading the data set & cleaning it
wineDT <- read.csv(here("data/winequality-red.csv"))
wine <- as_tibble(wineDT)
wine <- wine %>%
  clean_names()
colnames(wine)[9] <- "pH"
wine$quality <- as.factor(wine$quality)

```

## introduction
this report is a result of my exploration on [Red Wine Quality dataset](https://www.kaggle.com/datasets/uciml/red-wine-quality-cortez-et-al-2009), as a submission for the second assignment of Misk data science course.

to get back to the source repo [here](https://github.com/Misk-DSI/individual-assignment-2-Salma7577)

## Data

| Variable             | Description                                                                                                                                                                                           |
|---------------------|---------------------------------------------------|
| fixed_acidity        | most acids involved with wine or fixed or nonvolatile (do not evaporate readily)(g/ml)                                                                                                                |
| volatile_acidity     | the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste (g/ml                                                                                         |
| citric_acid          | found in small quantities, citric acid can add 'freshness' and flavor to wines(g/l)                                                                                                                   |
| residual_sugar       | the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet(g/l)                     |
| chlorides            | the amount of salt in the wine(g/l)                                                                                                                                                                   |
| free_sulfur_dioxide  | the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine(mg/l)                                 |
| total_sulfur_dioxide | amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine(mg/l) |
| density              | the density of water is close to that of water depending on the percent alcohol and sugar content(g/mL)                                                                                               |
| pH                   | describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale                                                                |
| sulphates            | a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant(mg/l)                                                                         |
| alcohol              | the percent alcohol content of the wine(%abv)                                                                                                                                                         |
| quality              | Quality of wine scale between 0 - 10 (target variable)                                                                          


|
```{r }
datatable(wine)
```




## EDA

firstly, we will take a look on the structure of the dataset to know what we are dealing with.

```{r }
str(wine)
```

we can see the name of the variables is slightly different(i.e. "janitor::clean_names()" has been used) and quality is now a factor(i.e "as.factor(wine\$quality)" used) instead of a an integer interval between 0-10, those changes has been made to make this EDA more convenient.


### Wine Quality distribution

```{r }
wine %>% 
  ggplot(aes(x=quality, fill=quality)) +
  geom_bar()
```

well this is interesting from the plot we notice that wine quality has a normal distribution most of the entities are between 5-6 with wines rated only inside the interval [3,8] instead of [0,10].

### Wine variables distribution

lets take a look at the other variables distribution

```{r  }
g1 <- wine %>% ggplot(aes(x=fixed_acidity)) + geom_histogram(bins = 20)
g2 <- wine %>% ggplot(aes(x=volatile_acidity)) + geom_histogram(bins = 20)
g3 <- wine %>% ggplot(aes(x=citric_acid)) + geom_histogram(bins = 20)
g4 <- wine %>% ggplot(aes(x=residual_sugar)) + geom_histogram(bins = 20)
g5 <- wine %>% ggplot(aes(x=chlorides)) + geom_histogram(bins = 20)
g6 <- wine %>% ggplot(aes(x=free_sulfur_dioxide)) + geom_histogram(bins = 20)
g7 <- wine %>% ggplot(aes(x=total_sulfur_dioxide)) + geom_histogram(bins = 20)
g8 <- wine %>% ggplot(aes(x=density)) + geom_histogram(bins = 20)
g9 <- wine %>% ggplot(aes(x=pH)) + geom_histogram(bins = 20)
g10 <- wine %>% ggplot(aes(x=sulphates)) + geom_histogram(bins = 20)
g11 <- wine %>% ggplot(aes(x=alcohol)) + geom_histogram(bins = 20)

grid.arrange(g1,g2,g3,g4,g5,g6,g7,g8,g9,g10,g11)
```

from the plots above we can see that the distribution of the variables are either normally distributed or right-skewed.

critc_aced plot shows a large number of wines with 0 critic_aced which actually tell us it is highly used in red wines.

Density and pH are normally distributed, which tells us that the red wines have a small difference in these two variables.

### quality variables relationship

#### quality & volatile_acidity

```{r }
 ggplot(wine,aes(x=quality, y= volatile_acidity, color = quality)) +
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot() 
```

it seems volatile_acidity has a negative effect on quality "downward trend".


#### quality & citric_acid

```{r }
ggplot(wine,aes(x=quality, y= citric_acid,color = quality)) +
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()  
```

it seems citric_acid has a positive effect on quality "upward trend".


#### quality & density

```{r }
ggplot(wine,aes(x=quality, y= density,color = quality)) +
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()  
```

it seems density has a negative effect on quality "downward trend".


#### quality & sulphates

```{r , echo=FALSE}
ggplot(wine,aes(x=quality, y= sulphates,color = quality)) + 
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()  
```

it seems sulphates has a positive effect on quality "upward trend".


#### quality & alcohol

```{r }
 ggplot(wine,aes(x=quality, y= alcohol,color = quality)) + 
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()
```

it seems alcohol has a positive effect on quality "upward trend".


#### other variables

```{r }
rel1 <- ggplot(wine,aes(x=quality, y= pH,color = quality)) + 
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()

rel2 <- ggplot(wine,aes(x=quality, y= total_sulfur_dioxide,color = quality)) + 
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot()

rel3 <- ggplot(wine,aes(x=quality, y= free_sulfur_dioxide, color = quality)) + 
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot() 

rel4 <- ggplot(wine,aes(x=quality, y= chlorides, color = quality)) +
   geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot() 

rel5 <- ggplot(wine,aes(x=quality, y= residual_sugar, color = quality)) +
  geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
  geom_boxplot() 

rel6 <- ggplot(wine,aes(x=quality, y= fixed_acidity, color = quality)) + 
   geom_jitter(width = 0.15, shape = 16, alpha = 0.75) +
   geom_boxplot() 


grid.arrange(rel1,rel2,rel3,rel4,rel5,rel6)

```

from the above plot we can see that these variables seems to have no effect on the quality of the red wine.


## Conclusion

from the EDA above we discovered a few interesting thing such as:

- quality ratings are only between 3 and 8 which indicates that there might be a systematic error or any kind of unbalance of those rating.

- quality ratings has a normal distribution along with some of the wines attributes such as density and pH.

- their is a positive relationship between the quality of the red wine and citric_acid,sulphates and alcohol more of those usually mean more quality.

- their is a negative relationship between the quality of the red wine and volatile_acidity,density.

- pH,total_sulfur_dioxide,free_sulfur_dioxide,chlorides,residual_sugar and fixed_acidity seems to have no effect on the quailty of the red wine.